---
title: "lab 19 - unsupervised mini project"
author: "Karina Cardenas"
format: html
---

# Background 

To identify somatic mutations in a tumor, DNA from the tumor is sequenced and compared
to DNA from normal tissue in the same individual using variant calling algorithms.
Comparison of tumor sequences to those from normal tissue (rather than ‘the human genome’)
is important to ensure that the detected differences are not germline mutations.
To identify which of the somatic mutations leads to the production of aberrant proteins, the
location of the mutation in the genome is inspected to identify non-synonymous mutations
(i.e. those that fall into protein coding regions and change the encoded amino acid).

> Q1 What protein do these sequences correspond to? (Give both full gene/protein name and oﬀicial symbol).

```{r}
library(seqinr)

fasta <- read.fasta(file = "A16742606_mutant_seq.fa", 
                    seqtype = "DNA", 
                    as.string = TRUE)

fasta$wt_healthy[1]
fasta$mutant_tumor[1]
```

> Q2. What are the tumor specific mutations in this particular case ( e.g. A130V)?

```{r}
library(bio3d)

seqs<- read.fasta("A16742606_mutant_seq.fa")

seqs
```

```{r}
getwd()

wt_healthy_seq <- fasta[["wt_healthy"]]

wt_healthy_seq
```

```{r}
aa_seq <- paste(wt_healthy_seq, collapse = "")

aa_seq
```

```{r}
blast <- blast.pdb(aa_seq)
```

```{r}
head(blast)
```

```{r}
blast$hit.tbl
```

```{r}
blast$hit.tbl[5,4] 
```

