---
title: "Lab11"
author: "Karina Cardenas, A16742606"
format: html
toc: true
---

## Comparative structure analysis of Adenylate Kinase

Load up the packages we will need for analysis of protein structure sets 
```{r}
library(bio3d)
library(NGLVieweR)
```

> Q10. Which of the packages above is found only on BioConductor and not CRAN? 

package found only in BioConductor is `msa` 

> Q11. Which of the above packages is not found on BioConductor or CRAN?: 

> Q12. True or False? Functions from the pak package can be used to install packages from 

**TRUE**


We will analyze the ADK starting with a single ADK database acession code: "1ake_A"
```{r}
id <- "1ake_A"

aa <- get.seq("1ake_A")

aa
```

> Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 



Now we can search the PDB database to find all related entries. 
```{r}
blast <- blast.pdb(aa)
```

```{r}
attributes(blast)
```

```{r}
head(blast$hit.tbl)
```

```{r}
hits <- plot(blast)
```

Our "top hits" i.e the most similar entries in the database are
```{r}
hits$pdb.id
```

compress all ADK hits
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```
Alignment: Align and superimpose these structures
```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```
sidenote: 
```{r}
library(bio3dview)
view.pdbs(pdbs)
```

This is better but still difficult to see what is similar and different in all these structures or indeed learn much about how this family works. 

Let's try PCA: 
```{r}
pc <- pca(pdbs)
plot(pc)
```

```{r}
plot(pc, pc.axes = 1:2)
```
```{r}
view.pca(pc)
```

Write a PDB "trajectory" for mol-star
```{r}
mktrj(pc,file = "pca_results.pdb")
```

