---
title: "lab12"
author: "Karina Cardenas, A16742606"
date: "May 11, 2025"
format: pdf
toc: true
---

## Section 1: Gene proprtions in population 

Finding the Proportion of Homozygous G|G genes in AMR,MXL descent 
```{r}
GG_proportion <- read.csv("sample_genotypes.csv")

table(GG_proportion$Genotype..forward.strand.)
```

```{r}
table(GG_proportion$Genotype..forward.strand.) / nrow(GG_proportion) * 100
```


Now lets look at a different population, I picked GBR.
```{r}
GBR_pop <- read.csv("GBR_pop.csv")

head(GBR_pop)
```


Find the proportion of G|G
```{r}
round(table(GBR_pop$Genotype..forward.strand.)/nrow(GBR_pop) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population. 


## Section 4: plotting 

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.


The file below contains the Genotype expression results of one individual male with childhood asthma. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.
```{r}
expression <- read.table("Expression_genotype.txt")

head(expression)
```

> Q.  how many samples do we have ? 

```{r}
nrow(expression)
```

> Q. sample size of genotype expression? 

```{r}
table(expression$geno)
```

> Q. Median Expression of genotypes? 

```{r}
exp_sum <- tapply(expression$exp, expression$geno, median, na.rm = TRUE)

exp_sum
```


Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?
```{r}
library(ggplot2)
```

```{r}
ggplot(expression) + 
  aes(x = geno, y = exp, fill = geno) + 
  geom_boxplot(notch = TRUE) + 
  geom_jitter(width = 0.2, size = 1, alpha = 0.2) + 
  labs( x = "Genotype", y = "Expression")
```

